---
title: "Assignment 3"
output: html_document
---

```{r include=FALSE}
setwd("~/Documents/GitHub/SFBI/Assignment 3/working folder")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::install_github("jamgreen/lehdr")

library(lehdr)
library(tidyverse)
library(tidytransit)
library(remotes)
library(sf)
library(censusapi)
library(tigris)
library(leaflet)
library(readxl)
library(mapboxapi)
library(mapview)

(tigris_use_cache <- FALSE)

Sys.setenv(CENSUS_KEY = "c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r eval=FALSE, include=FALSE}
ca_od_read <- read_csv("/Volumes/GoogleDrive-101808300713557119266/Shared drives/SFBI/Data Library/LODES/ca_od_main_JT01_2019.csv.gz")
```


```{r eval=FALSE, include=FALSE}
zctas <- zctas()
blocks <- blocks("CA")
```

```{r eval=FALSE, include=FALSE}
zip_94127 <- zctas %>%
  filter(GEOID10 == "94127")

zip_blocks_94127 <- blocks %>%
  st_centroid() %>%
  .[zip_94127, ]
```

```{r eval=FALSE, include=FALSE}
zip_od_94127 <- ca_od_read %>%
  filter(
    h_geocode %in% zip_blocks_94127$GEOID10 |
      w_geocode %in% zip_blocks_94127$GEOID10
  )
```

```{r eval=FALSE, include=FALSE}
saveRDS(zip_94127, "zip_94127.rds")
saveRDS(zip_blocks_94127, "zip_blocks_94127.rds")
saveRDS(zip_od_94127, "zip_od_94127.rds")
```

```{r eval=FALSE, include=FALSE}
zip_blocks_94127 <- readRDS("zip_blocks_94127")

full_zip_od_94127 <- 2013:2019 %>%
  map_dfr(function(year) {
    print(year)

    temp <- read_csv(paste0("/Volumes/GoogleDrive-101808300713557119266/Shared drives/SFBI/Data Library/LODES/ca_od_main_JT01_", year, ".csv.gz")) %>%
      filter(
        h_geocode %in% zip_blocks_94127$GEOID10 |
          w_geocode %in% zip_blocks_94127$GEOID10
      ) %>%
      mutate(year = year)

    saveRDS(temp, paste0("temp_od_", year, ".rds"))

    return(temp)
  })
saveRDS(full_zip_od_94127, "full_zip_od_94127.rds")
```

```{r include=FALSE}
full_zip_od_94127 <- readRDS("full_zip_od_94127.rds")
zip_blocks_94127 <- readRDS("zip_blocks_94127.rds")

full_zip_od_94127_clean <- full_zip_od_94127 %>%
  select(-createdate) %>%
  filter(!(
    h_geocode %in% zip_blocks_94127$GEOID10 &
      w_geocode %in% zip_blocks_94127$GEOID10
  )) %>%
  mutate(
    direction = ifelse(
      h_geocode %in% zip_blocks_94127$GEOID10,
      "outbound",
      "inbound"
    )
  )
saveRDS(full_zip_od_94127_clean, "full_zip_od_94127_clean.rds")
```

```{r include=FALSE}
mapbox_call_dataset_94127 <- full_zip_od_94127_clean %>%
  mutate(
    start = pmin(h_geocode, w_geocode),
    end = pmax(h_geocode, w_geocode),
    origin = ifelse( # this is an interesting step that I should review a little bit more
      direction == "inbound",
      h_geocode,
      w_geocode
    ),
    cbg = origin %>% substr(1, 12), # this is moving up to cbg
    tract = origin %>% substr(1, 11)
  ) %>%
  filter(!duplicated(paste(start, end))) %>%
  filter(!duplicated(paste(origin))) %>%
  filter(!duplicated(paste(tract)))
```

```{r include=FALSE}
tract_job_counts <- full_zip_od_94127_clean %>%
  mutate(
    start = pmin(h_geocode, w_geocode),
    end = pmax(h_geocode, w_geocode),
    origin = ifelse( # this is an interesting step that I should review a little bit more
      direction == "inbound",
      h_geocode,
      w_geocode
    ),
    cbg = origin %>% substr(1, 12), # this is moving up to cbg
    tract = origin %>% substr(1, 11)
  ) %>%
  group_by(tract) %>%
  summarise(n_jobs = sum(S000))
```


```{r eval=FALSE, include=FALSE}
zip_94127 <- readRDS("zip_94127.rds")

ca_tracts <- tracts("CA")

zip_od_origin_94127 <-
  mapbox_call_dataset_94127 %>%
  select(tract) %>%
  left_join(ca_tracts %>% select(tract = GEOID)) %>%
  st_as_sf() %>%
  st_centroid() %>%
  st_coordinates()

zip_od_destination_94127 <-
  zip_94127 %>%
  st_centroid() %>%
  st_coordinates()
```

```{r eval=FALSE, include=FALSE}
zip_od_route_94127 <-
  1:ceiling(nrow(zip_od_origin_94127) / 1000) %>%
  map_dfr(function(y) {
    print(y)

    temp <- (y * 1000 - 999):pmin(y * 1000, nrow(zip_od_origin_94127)) %>%
      map_dfr(function(x) {
        tryCatch(
          mb_directions(
            origin = zip_od_origin_94127[x, ],
            destination = zip_od_destination_94127,
            profile = "driving-traffic"
          ) %>%
            mutate(id = x),
          error = function(e) {
            data.frame(id = x)
          }
        )
      }) %>%
      st_as_sf()

    saveRDS(temp, paste0("temp", y, ".rds"))

    return(temp)
  })
```

```{r}

```


```{r include=FALSE}
zip_od_route_94127 <- 1:4 %>%
  map_df(function(x) {
    readRDS(paste0("temp", x, ".rds"))
  })


full_zip_od_routed_94127 <- mapbox_call_dataset_94127 %>%
  cbind(zip_od_route_94127)

full_zip_od_final_94127 <- full_zip_od_94127_clean %>%
  mutate(
    origin = ifelse(
      direction == "inbound",
      h_geocode,
      w_geocode
    ),
    tract = substr(origin, 1, 11)
  ) %>%
  left_join(
    full_zip_od_routed_94127 %>%
      select(cbg, duration, distance, tract),
    by = "tract"
  ) %>% 
  filter(duration < 160) %>% # looking at only trips under 3 hours) %>%
  mutate(
    visits = if_else(duration < 100,
      S000 * 261, # annual work days
      S000 * 130,
    ),
    # number of work days annually,
    vmt = visits * distance*2 # 2 trips a day per job
  )

saveRDS(full_zip_od_final_94127, ("full_zip_od_final_94127.rds"))
```

The zip code that I live in is largely residential.  Though technically in San Francisco,
my neighborhood primarily consists of single family homes with a few commercial stretches
of groceries or other stores.  In calculating the GHG emissions for this zip code,
I would expect that the majority of emissions are generated from commuters leaving the
neighborhood on a daily basis.  In my transportation GHG calculation, I also 
include the trips that commute into the zip code.  This is where I start my 
analysis: I imported Census LODES job data to filter down all the jobs that fit these 
conditions (i.e. commuting in or out of my zip code, 94127).  I looked at home and
work geocodes and did not include any pairs where both job and home were in 94127.
The reasoning being that these trips contributed a negligible amount of Vehicle Miles
Traveled (VMTs) and complicated the analysis.

I generated trip geometries from the MapboxAPI and plotted all commute times here:

```{r echo=FALSE}


leaflet() %>%
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>%
  addPolylines(
    data = zip_od_route_94127 %>%
      filter(
        duration < 120 # looking at only trips under 2 hours
      ) 
  )
```


Initially this map included trips across the entire state; this is likely due to
residents in this zip code having business registered in other parts of the state.
Had my I included zips nationwide, I would expect to have trips coming from across
the country.  To control for these outlier trips (which would significantly skew the
total VMTs), I am only considering trip times under 2 hours.  This is roughly four
times the mean commute time in the bay area commute as reported by 
[CBS SF Bay Area](https://sanfrancisco.cbslocal.com/2020/05/06/coronavirus-covid-19-era-commute-times-could-double-if-bay-area-workers-ditch-transit-for-cars/)

In my actual totaling of VMTs, I throw out the most extreme trips entirely.  What 
I classify as "long commutes" (those > 100 minutes one way), I assume 
visits happen less frequently than those under 100 minutes.  I calculate those trip 
totals by multiplying the trip miles by 261 days for an annual count.  For the trips
that they are not happening daily, I assume that they happen half as frequently.
In the future, this could be estimated more robustly by using NHTS median commute
time data.  Using standard deviations might be a less arbitrary way to assign
trip counts, but these numbers are reasonable enough for the purpose of this 
analysis.

Since there are different emission rates associated with different vehicles, I have
to consider the types of vehicles and fuels used in trips for different years. To 
incorporate this, I use annual EMFAC data to weight emissions associated with 
the VMTs in aggregate each year.  


```{r include=FALSE}
emfac <-
  read_csv("emfac.csv", skip = 8) %>%
  transmute(
    Year = `Calendar Year`,
    Category = `Vehicle Category`,
    Fuel_Type = Fuel,
    Percent_Trips = Trips / sum(Trips),
    Percent_Miles = `Total VMT` / sum(`Total VMT`),
    `MTCO2_Running_Exhaust` = CO2_RUNEX / `Total VMT`,
    `MTCO2_Start_Exhaust` = CO2_STREX / Trips
  )

trips_94127_ghg <- full_zip_od_final_94127 %>%
  group_by(year) %>%
  summarise(
    visits = sum(visits),
    vmt = sum(vmt)
  ) %>%
  left_join(emfac, by = c("year" = "Year")) %>%
  mutate(
    trips = Percent_Trips * visits,
    vmt = Percent_Miles * vmt,
    ghg_start = trips * MTCO2_Start_Exhaust * 2,
    ghg_running = vmt * MTCO2_Running_Exhaust,
    ghg = ghg_start + ghg_running
  ) %>%
  group_by(year) %>%
  summarise(across(starts_with("ghg"), sum))
```

```{r echo=FALSE}
  trips_94127_ghg %>% ggplot(
    aes(
      x = year,
      y = ghg
    )
  ) +
  geom_line(stat = "identity") +
  labs(x = "Year", y = "Tons of CO2", title = "94127 Commute Emissions, 2013 to 2019") +
  scale_x_continuous(breaks = 2013:2019) +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal()
```


This plot gives some indication that commute emissions associated with the 94127 
zip code have started to decrease over the last 3 years.  This could be a result
of changes in fuel types for these trips.  Electric cars, for instance, do not
have any associated emissions in this data set; the 94127 zip code likely has a
high adoption rate of EVs.

## Building Emissions

Useful to this analysis of GHGs from 94127 is to consider the  emissions from 
buildings as well.  Since 2013, PG&E publishes an estimated conversion factor from
kWh to tons of CO2e. Using these factors and the reported kBTU and kWh used in 
94127, I can total emissions from different building types (commercial
and residential), and energy usage (gas and electric).  One point to call out is
that the factor for 2019 that PG&E reports is two orders of magnitude smaller 
than the previous years.  This is to do with a change in how they are required 
to report emissions with Community Choice Aggregators and is why there is such a 
sharp decline in emissions from electricity in 2019. 

In the following graph, I plotted the commute emissions with building emissions
from each year.

```{r eval=FALSE, include=FALSE}
pge_elec_emissions_factor <-
  data.frame(
    year = c(2013:2019),
    factor = c(427, 435, 405, 294, 210, 206, 2.68)
  )

pge_data <-
  2013:2019 %>%
  map_dfr(function(yr) {
    factor <-
      pge_elec_emissions_factor %>%
      filter(year == yr) %>%
      pull(factor)
    1:4 %>%
      map_dfr(function(quarter) {
        c("Electric", "Gas") %>%
          map_dfr(function(type) {
            filename <-
              paste0(
                "/Volumes/GoogleDrive-101808300713557119266/Shared drives/SFBI/Data Library/PG&E/PGE_",
                yr,
                "_Q",
                quarter,
                "_",
                type,
                "UsageByZip.csv"
              )
            temp <- read_csv(filename, show_col_types = FALSE)
            if (yr == 2017 & quarter == 4) {
              temp <-
                temp %>%
                filter(MONTH != 9)
            }
            temp <-
              temp %>%
              rename_all(toupper) %>%
              mutate(
                TOTALKBTU = ifelse(
                  substr(CUSTOMERCLASS, 1, 1) == "E",
                  TOTALKWH * 3.412,
                  TOTALTHM * 99.976
                ),
                TOTALTCO2E = ifelse(
                  substr(CUSTOMERCLASS, 1, 1) == "E",
                  TOTALKWH / 1000 * factor * 0.000453592,
                  TOTALTHM * 0.00531
                )
              ) %>%
              select(
                ZIPCODE,
                YEAR,
                MONTH,
                CUSTOMERCLASS,
                TOTALKBTU,
                TOTALTCO2E,
                TOTALCUSTOMERS
              )
          })
      })
  })
```

```{r eval=FALSE, include=FALSE}
us_zips <-
  zctas(cb = T, progress_bar = F)

sf_zips <-
  us_zips %>%
  st_centroid() %>%
  .[counties("CA", cb = T, progress_bar = F) %>% filter(NAME == "San Francisco"), ] %>%
  st_drop_geometry() %>%
  left_join(us_zips %>% select(GEOID10)) %>%
  st_as_sf() %>%
  st_transform(4326)
```

```{r eval=FALSE, include=FALSE}
sf_pge_data <-
  pge_data %>%
  filter(ZIPCODE == "94127") %>%
  filter(CUSTOMERCLASS %in% c(
    "Elec- Commercial",
    "Elec- Residential",
    "Gas- Commercial",
    "Gas- Residential"
  )) %>%
  separate(CUSTOMERCLASS, c("Energy_Type", "Customer_Type"), sep = "- ") %>%
  group_by(ZIPCODE, Energy_Type, Customer_Type, YEAR) %>%
  summarize(
    TOTALKBTU = sum(TOTALKBTU, na.rm = T),
    TOTALTCO2E = sum(TOTALTCO2E, na.rm = T),
    TOTALCUSTOMERS = mean(TOTALCUSTOMERS, na.rm = T)
  ) %>%
  group_by(Customer_Type, Energy_Type, YEAR) %>%
  summarize(across(
    c(TOTALKBTU, TOTALTCO2E, TOTALCUSTOMERS),
    ~ sum(., na.rm = T)
  ))
```

```{r echo=FALSE}
sf_pge_data <- readRDS("sf_pge_data.rds")

combined_building_transit <- sf_pge_data %>% 
  ungroup() %>% 
  select(YEAR, TOTALTCO2E) %>% 
  group_by(YEAR) %>% 
  summarise(
    TOTALTCO2E = sum(TOTALTCO2E)
  ) %>% 
  mutate(
    source = "Building"
  ) %>% 
  rbind(
    trips_94127_ghg %>% 
      rename(YEAR = year) %>% 
      group_by(YEAR) %>% 
      summarise(
      TOTALTCO2E = sum(ghg)
      ) %>% 
      mutate(
        source = "Commute"
      )
  )
  
combined_building_transit %>%
  ggplot(
    aes(
      x = YEAR,
      y = TOTALTCO2E
    )
  ) +
  geom_bar(stat = "identity", aes(fill = source), position = "stack") +
  labs(x = "Year", y = "Tons of CO2e", title = "94127 Annual Emissions, 2013 to 2019") +
  scale_fill_discrete(name = "Emission Source", labels = c("Building", "Commute")) + 
  scale_x_continuous(breaks = 2013:2019) +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal()
```

It is somewhat difficult to see much change in the building emissions year to year.
The following plot zooms in on what types of energy use is contributing most to
CO2e building emissions.  Clear here is that gas usage is a much bigger emitter 
than electricity, which has contributed progressively less CO2e.  The near 0 tons 
of CO2e reported in 2019 is result of the above mentioned change in reporting. 
Nonetheless, in the years prior, electricity has consistently been contributing 
less CO2e.

```{r echo=FALSE}
ggplot(
  sf_pge_data,
  aes(
    x = as.factor(YEAR),
    y = TOTALTCO2E
  )
) +
  geom_bar(stat = "identity", aes(fill = Energy_Type), position = "dodge") +
  labs(x = "Year", y = "Tons of CO2e", title = "94127 Building Emissions, 2013 to 2019") +
  scale_fill_discrete(name = "Energy Type", labels = c("Electricity", "Gas")) +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal()
```

The trend in electricity emissions is promising, especially when considering that 
there has been little change in the proportion of energy usage between gas and 
electricity.  This is to say that the decrease in emissions from electricity is 
not associated with a decrease in electricty usage, as clearly seen in the plot
below. 


```{r echo=FALSE}
ggplot(
  sf_pge_data,
  aes(
    x = as.factor(YEAR),
    y = TOTALKBTU / 1000000
  )
) +
  geom_bar(stat = "identity", aes(fill = Energy_Type), position = "dodge") +
  labs(x = "Year", y = "GBTU", title = "94127 Annual Energy Usage, 2013 to 2019") +
  scale_fill_discrete(name = "Energy Type", labels = c("Electricity", "Gas")) +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal()
```

A deeper investigation into building type can reveal other interesting findings.
Since 94127 is a largely residential neighborhood, it is reasonable to expect that
residential buildings would contribute more to the ultimate GHG totals.  A plot of
the type of energy by the type of buildings shows that residential gas is far and
away the largest contributor of GHGs in the zipcode.


```{r echo=FALSE}
sf_pge_data %>%
  ggplot(
    aes(
      x = YEAR,
      y = TOTALTCO2E
    )
  ) +
  geom_line(aes(color = Energy_Type, linetype = Customer_Type)) +
  labs(x = "Year", y = "tCO2e", title = "94127 Annual Emissions By Type, 2013 to 2019") +
  scale_color_discrete(name = "Energy Type", labels = c("Electricity", "Gas"))+
  scale_x_continuous(breaks = 2013:2019) +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal()
```


Within this zip code, electrifying homes would appear to have an enormous effect 
on the overall associate emissions.  As there are not many commercial properties,
neither the commercial gas or electricity emissions are proportionately very high.

A normalized energy usage could potentially give a better sense of which buildings
are operating more energy intensively.  Below, I normalize commercial energy use 
by the number of jobs within 94127 and residential energy use by the reported 
total of 94127 customers from PG&E. I also want to normalize to the difference 
in temperatures year to year, as a year with a record heat wave would likely have 
greater electricity usage for cooling than other years. I divided electricity 
usage by the heating degree days (HDDs), as those days require more AC loads for 
cooling,and gas usage by cooling degree days (CDDs) which would drive a higher 
demand for gas heating in most buildings.

```{r include=FALSE}
HDD_94127 <- read_csv("HDD_94127.csv", skip = 8) %>%
  select(year, "CanESM2 (Average)") %>%
  filter(
    year == 2013:2019
  ) %>%
  group_by(year) %>%
  transmute("HDD_average" = `CanESM2 (Average)`)

CDD_94127 <- read_csv("CDD_94127.csv", skip = 8) %>%
  select(year, "CanESM2 (Average)") %>%
  filter(
    year == 2013:2019
  ) %>%
  group_by(year) %>%
  transmute("CDD_average" = `CanESM2 (Average)`)
```


```{r include=FALSE}
full_zip_od_final_94127 <- readRDS("full_zip_od_final_94127.rds")
zip_blocks_94127 <- readRDS("zip_blocks_94127.rds")
jobs_count <- full_zip_od_final_94127 %>%
  filter(w_geocode %in% zip_blocks_94127$GEOID10) %>% 
  group_by(year) %>%
  summarise(
    total_jobs = sum(S000)
  )
```

```{r message=TRUE, include=FALSE}
normalized_building_emissions <- sf_pge_data %>%
  left_join(jobs_count %>% rename(YEAR = year)) %>%
  left_join(CDD_94127 %>% rename(YEAR = year)) %>%
  left_join(HDD_94127 %>% rename(YEAR = year)) %>%
  ungroup() %>%
  mutate(
    intensity_by_pop_hdd = if_else(
      Customer_Type == "Residential" & Energy_Type == "Elec",
      (TOTALTCO2E / TOTALCUSTOMERS) / HDD_average,
      NA_real_
    ),
    intensity_by_pop_cdd = if_else(
      Customer_Type == "Residential" & Energy_Type == "Gas",
      (TOTALTCO2E / TOTALCUSTOMERS) / CDD_average,
      NA_real_
    ),
    intensity_by_job_hdd = if_else(
      Customer_Type == "Commercial" & Energy_Type == "Elec",
      (TOTALTCO2E / total_jobs) / HDD_average,
      NA_real_
    ),
    intensity_by_job_cdd = if_else(
      Customer_Type == "Commercial" & Energy_Type == "Gas",
      (TOTALTCO2E / total_jobs) / CDD_average,
      NA_real_
    )
  )

```

```{r echo=FALSE}
normalized_building_emissions %>%
  mutate(
    line_group = rep(c("residential_elec", "residential_gas", "commercial_elec", "commercial_gas"), each = 7),
    emissions_intensity = coalesce(intensity_by_pop_hdd, intensity_by_pop_cdd, intensity_by_job_hdd, intensity_by_job_cdd)
  ) %>%
  ggplot(
    aes(
      x = YEAR,
      y = emissions_intensity
    )
  ) +
  geom_line(aes(color = line_group)) +
  labs(
    x = "Year",
    y = "tCO2e/Population/Degree Differential",
    title = "94127 Annual Energy Usage Normalized by Intensity, 2013 to 2019"
  ) +
  theme_minimal()
```

The normalized plot of different building types and energy usage indicates that
commercial electricity may actually be contributing a disproportionate amount of
CO2e emissions.  In considering where the greatest opportunity lies in reducing 
carbon emissions however, this analysis perhaps suggests that a reduction in 
commercial gas (for this specific zip code) could easily be made less energy
intensive, even though they contribute relatively little to the total CO2e emissions.
One caveat is that PG&E is missing data for much of the commercial electricity
reporting, so this plot should be taken with a grain of salt.  Some conclusions
that can be confidently drawn however, is that building energy from gas is a 
hugely significant contributor of CO2e emissions in 94127 and ought to be a target
for reducing GHG emissions.